Compare scANVI and SingleR annotations

### 2025-08-14 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/scanvi-singler-comparison.Rmd
add section comparing bad libraries

merged_dir <- file.path(repository_base, "data", "current", "results", "merge-sce", "SCPCP000004")
module_dir <- file.path(repository_base, "analyses", "cell-type-neuroblastoma-04")
ref_dir <- file.path(module_dir, "references")
results_dir <- file.path(module_dir, "results")
singler_dir <- file.path(results_dir, "singler")
scanvi_dir <- file.path(results_dir, "scanvi")
```

@@

# label mapping
nbatlas_label_map_file <- file.path(ref_dir, "nbatlas-label-map.tsv")
```

@@

nbatlas_palette_df <- readr::read_tsv(nbatlas_palette_file)
nbatlas_celltype_pal <- nbatlas_palette_df$hex_color
names(nbatlas_celltype_pal) <- nbatlas_palette_df$NBAtlas_label
```

@@

Across libraries, what fraction of `SingleR` and `scANVI` labels exactly agree?
Note that this analysis does not consider similar labels (e.g. two types of T cells) but only does a quick look at identical labels.
We'll also highlight the PDX samples here to see if there was any associated bias.

> I think it would be helpful to get a table in here that shows you % of each library that has exact agreement between SingleR and scANVI, % of each library where cells are in the same family (e.g., the same cell is labeled as two different T cell types), and % of each library that disagrees and is not in the same family.

In this case, we consider these families specifically:

* T cells: `T cell`, `Treg`, `CD4+ T cell`, `CD8+ T cell`
* Natural killer cells: `NK cell`, `Circulating NK cell`, `Resident NK cell`, `TOX2+/KIT+ NK cell`
* Conventional dendritic cells (cDC): `cDC`, `cDC1`, `cDC2/DC3`, `Migratory cDC`
* Monocytes: `Monocyte`, `Classical monocyte`, `Patrolling monocyte`
* Myeloid: `Neutrophil` and `Macrophage`

@@

```{r}
# combine into a single table for display & viz
frac_df <- identical_agree_df |>
  dplyr::left_join(family_agree_df) |>
  dplyr::left_join(disagree_df) |>
  # round to 4 decimals
  dplyr::mutate(dplyr::across(contains("frac_"), round, 4)) |>
  dplyr::select(
    library_id, 
    library_size, 
    frac_identical, 
    frac_same_family, 
    frac_disagree
  ) |>
  dplyr::arrange(library_size)

@@

```{r}
# add PDX column for plotting
frac_df <- celltype_df |> 
  dplyr::select(library_id, is_xenograft) |>
  unique() |>
  dplyr::right_join(frac_df)

ggplot(frac_df) + 
  aes(x = library_size, y = frac_same_family) + 
  geom_point(aes(color = is_xenograft)) + 
  geom_smooth(method = "lm") +
  labs(
    x = "Library size",
    y = "Fraction of SingleR/scANVI labels from same family"
  ) +
  # stop going above 1
  scale_y_continuous(limits = c(0, 1))
```

Most cells have agreeing annotations in the majority of libraries.
There appears to be a slight trend (not signficant) that larger libraries have more agreement between methods, but this also seems driven by the libraries with low correspondence.

@@

>>>>

data_dir <- file.path(repository_base, "data", "current", "SCPCP000004")
merged_dir <- file.path(repository_base, "data", "current", "results", "merge-sce", "SCPCP000004")
module_dir <- file.path(repository_base, "analyses", "cell-type-neuroblastoma-04")
ref_dir <- file.path(module_dir, "references")
results_dir <- file.path(module_dir, "results")
singler_dir <- file.path(results_dir, "singler")
scanvi_dir <- file.path(results_dir, "scanvi")
```

@@

# label mapping
nbatlas_label_map_file <- file.path(ref_dir, "nbatlas-label-map.tsv")

# sample metadata
sample_metadata_file <- file.path(data_dir, "single_cell_metadata.tsv")
```

@@

nbatlas_palette_df <- readr::read_tsv(nbatlas_palette_file)
nbatlas_celltype_pal <- nbatlas_palette_df$hex_color
names(nbatlas_celltype_pal) <- nbatlas_palette_df$NBAtlas_label

# sample metadata
sample_metadata <- readr::read_tsv(sample_metadata_file)
```

@@

Across libraries, what fraction of `SingleR` and `scANVI` labels exactly agree?
For this, we'll compare among:

* Fraction of identical labels
* Fraction of labels that agree at least on the level of their cell type "family," where we define these families based on groups of NBAtlas annotations:
  * T cells: `T cell`, `Treg`, `CD4+ T cell`, `CD8+ T cell`
  * Natural killer cells: `NK cell`, `Circulating NK cell`, `Resident NK cell`, `TOX2+/KIT+ NK cell`
  * Conventional dendritic cells (cDC): `cDC`, `cDC1`, `cDC2/DC3`, `Migratory cDC`
  * Monocytes: `Monocyte`, `Classical monocyte`, `Patrolling monocyte`
  * Myeloid: `Neutrophil` and `Macrophage`
* Fraction of labels that entirely disagree

The table below shows these fractions for each library along with the library size.
We also visualize the relationship between library size and family-based agreement.

@@

```{r, message = FALSE}
# combine into a single table for display & viz
frac_df <- identical_agree_df |>
  dplyr::left_join(family_agree_df) |>
  dplyr::left_join(disagree_df) |>
  # round to 4 decimals
  dplyr::mutate(dplyr::across(contains("frac_"), \(x) round(x, 4))) |>
  dplyr::select(
    library_id, 
    library_size, 
    frac_identical, 
    frac_same_family, 
    frac_disagree
  ) |>
  dplyr::arrange(library_size)

@@

```{r, message = FALSE, fig.width = 5, fig.height = 3}
# add PDX column for plotting
frac_df <- celltype_df |> 
  dplyr::select(library_id, is_xenograft) |>
  unique() |>
  dplyr::right_join(frac_df)

ggplot(frac_df) + 
  aes(x = library_size, y = frac_same_family) + 
  geom_point(aes(color = is_xenograft)) + 
  geom_smooth(method = "lm") +
  labs(
    x = "Library size",
    y = "Fraction of SingleR/scANVI labels\nfrom the same family"
  ) +
  # stop going above 1
  scale_y_continuous(limits = c(0, 1))
```

Most cells have agreeing annotations in the majority of libraries.
Considering closely-related cells in the same family usually provides a couple more cells to be annotated, but not a large percentage.
There appears to be a slight trend (not significant) that larger libraries have more agreement between methods, but this also seems driven by the libraries with low correspondence.

@@

Here is the sample metadata for these four samples:

```{r}
disagree_ids <- frac_df |>
  dplyr::filter(frac_disagree >= 0.5) |>
  dplyr::pull(library_id)

disagree_metadata <- sample_metadata |>
  dplyr::filter(scpca_library_id %in% disagree_ids)
```

Looking closely here, we can see that these samples were in fact derived from two participants where one sample is a xenograft and the other is tissue.

```{r}
disagree_metadata |>
  dplyr::select(scpca_library_id, sample_type, participant_id) |>
  dplyr::arrange(participant_id)
```

These participant IDs are not associated with any other samples:

```{r}
sample_metadata |>
  dplyr::filter(participant_id %in% disagree_metadata$participant_id) |>
  dplyr::filter(!(scpca_library_id) %in% disagree_ids)
```

It's possible that patient-specific biology explains the discrepancies we see for these samples' cell types.

We'll look more closely at their disagreements as well with heatmaps:

```{r, fig.height = 8, fig.width = 8} 
disagree_ids |>
  # walk to only print the heatmaps once
  purrr::walk(
    \(id) {
      library_df <- celltype_wide_df |>
        dplyr::filter(library_id == id)

     frac_disagree <- frac_df |>
       dplyr::filter(library_id == id) |>
       dplyr::pull(frac_disagree)

     heatmap_title <- glue::glue("{id} (fraction disagree = {frac_disagree})")

      make_jaccard_heatmap(
        library_df,
        "scanvi",
        "singler",
        glue::glue("{heatmap_title}\n\nscANVI label"),
        "SingleR label"
      )
  })
```

For these libraries, the main source of disagreement is `scANVI` applying more homogeneous labeling of `Neuroendocrine` compared to `SingleR` which is assigning more varied cell types.
Note that the two libraries where `scANVI` is _only_ detecting `Neuroendocrine` cells are the PDX libraries, for which we expect lower cell type diversity in general.
It's not unlikely that a lot of what `SingleR` has labeled `Stromal other` or `Schwann` may be tumor.
